output:
    html_document:
        toc: true
        theme: cosmo
author: Adrian Martin, Rochester NY

## ZAP Feedback

I am writing this page as feedback to the Rochester City Planning Commission on the Zoning Alignment Project release of 2026. Feedback for the [June 10th meeting](https://rochesterzap.com/2026/05/13/city-planning-commission-and-rochester-environmental-commission-meeting-dates-announced/) is due June 9th, 4PM. This will be emailed to the commission, but also posted on github for public access and reproducibility.

My feedback looks at a few areas that I am concerned about:
- Downzoning specific neighborhoods
- Large numbers of parcels remaining non-compliant with zoning that does not reflect the current or desired (as per the Roc2034 comprehensive plan) character of that neighborhood

In addition to feedback on the ZAP plan, I also want to stress that I banged this analysis out in a few days because the city refused to proactively post the ZAP zoning file and required me to FOIL it (FOIL 2026-2176 filed Apr 20 2026). They then granted themselves an extension on that FOIL request (May 18 2026), despite the request requiring zero redaction or processing. I finally received the data on June 1st, 2026. This mirrored the process for releasing the 2023 ZAP zoning file (FOIL 2023-6693, filed Oct 29 2023, released Jan 26 2024, comment deadline was Jan 31 2024).
The city has a good open data portal, which I pull the current zoning file from. The city refused to provide any draft zoning file unless I filed a FOIL, and they declined to explain the reasoning behind their refusal. I know of no reason to prevent the city from proactively postoing the ZAP file to the open data portal back in April, when they released their plans. This whole process was extremely frustrating.

```{r setup, include=FALSE}

options(cache = TRUE)

library(tidyverse)
library(sf)
library(plotly)

```

```{r loaddata, message = FALSE, warning = FALSE}

# note: files are transformed to a projected coordinate system
# this is necessary for accurate spatial joins, calculating areas/distances, etc.
# leaflet requires lat/lon, so I transform to 4326 after the math is done

current_zoning <- read_sf("https://maps.cityofrochester.gov/server/rest/services/Open_Data/Zoning_Districts_Open_Data/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson") %>%
    rename(Zoning_current = LABEL) %>%
    st_transform(6541)

parcels_2022 <- read_sf("https://maps.cityofrochester.gov/server2/rest/services/Open_Data/TaxParcel2022/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson") %>% 
    st_transform(6541)

zap_2023 <- read_sf("DRAFT_ZAP_Districts_Sept2023/DRAFT_ZAP_DISTRICTS_Sept2023.shp") %>%
    rename(Zoning_2023 = NAME) %>%
    st_transform(6541)

zap_2026 <- read_sf("ZAP_Districts_final_draft_Apr26.shp") %>%
    rename(Zoning_2026 = NAME) %>%
    st_transform(6541)

# for plotting
ZAPwgs <- zap_2026 %>% st_transform(4326) %>%
  # don't plot the airport overlay
  filter(Zoning_2026 != "O-A")

```

```{r analysis}

zoningAreas <- current_zoning %>% 
  mutate(area = st_area(current_zoning)) %>%
  group_by(CATEGORY) %>%
  summarize(
    totalarea = sum(area),
    numDistricts = n()
  ) %>%
  st_drop_geometry() %>%
  mutate(acres= units::set_units(totalarea, "acres")) %>%
  arrange(desc(totalarea))

ZAPareas <- zap_2026 %>% 
  dplyr::filter(Zoning_2026 != "O-A") %>% #airport overlay not relevant
  mutate(area = st_area(zap_2026 %>% dplyr::filter(Zoning_2026 != "O-A"))) %>% 
  group_by(Zoning_2026) %>%
  summarize(
    totalarea = sum(area),
    numDistricts = n()
  ) %>%
  st_drop_geometry() %>%
  mutate(acres = units::set_units(totalarea, "acres")) %>%
  arrange(desc(totalarea))

parcelsJoin <- parcels_2022 %>% 
  mutate(parcelArea = st_area(.)) %>%
  st_join(zap_2026, st_overlaps, largest = TRUE) %>%
  #st_join(st_make_valid(ZAPwgs), st_overlaps) %>%
  st_join(current_zoning, st_overlaps, largest = TRUE)

parcelsSummed <- parcelsJoin %>% 
  group_by(CATEGORY, Zoning_2026) %>% 
  st_drop_geometry() %>%
  summarize(
    parcelCount = n(), 
    parcelArea = sum(parcelArea),
    parcelValue = sum(CURRENT_TOTAL_VALUE, na.rm = TRUE)) # 59 parcels with NA value

```

### Changes in total {.tabset}

```{r countsSankey}

nodeswithnums <- tibble(
  nodename = c(unique(parcelsSummed$CATEGORY),
               unique(parcelsSummed$Zoning_2026)),
  type = c(rep("Current zoning", times = length(unique(parcelsSummed$CATEGORY))),
           rep("ZAP_2026", times = length(unique(parcelsSummed$Zoning_2026))))
) %>%
  mutate(nodenumber = row_number() - 1)

flowvals <- parcelsSummed %>%
  left_join(
    nodeswithnums %>% 
      filter(type == "Current zoning") %>%
      dplyr::select(-type),
    by = c("CATEGORY" = "nodename")
  ) %>% 
  rename("sourcenode" = "nodenumber") %>%
  left_join(
    nodeswithnums %>%
      filter(type == "ZAP_2026") %>%
      dplyr::select(-type),
    by = c("Zoning_2026" = "nodename")
  ) %>%
  rename("targetnode" = "nodenumber")

countSankey <- plot_ly(
  type = "sankey",
  domain = list(c(x = c(0, 1), y = c(0, 1))),
  orientation = "h",
  arrangement = "snap",
  node = list(
    label = nodeswithnums$nodename,
    pad = 15, 
    thickness = 15
  ),
  link = list(
    source = flowvals$sourcenode,
    target = flowvals$targetnode,
    value = flowvals$parcelCount
  )
) %>%
  layout(
    title = "Counts of parcels moving from current zoning to 2026 ZAP",
    xaxis = list(showgrid = F, zeroline = F, title = "", showticklabels = FALSE),
    yaxis = list(showgrid = F, zeroline = F, title = "", showticklabels = FALSE)
  )

areaSankey <- plot_ly(
  type = "sankey",
  domain = list(c(x = c(0, 1), y = c(0, 1))),
  orientation = "h",
  arrangement = "snap",
  node = list(
    label = nodeswithnums$nodename,
    pad = 15, 
    thickness = 15
  ),
  link = list(
    source = flowvals$sourcenode,
    target = flowvals$targetnode,
    value = units::set_units(flowvals$parcelArea, "acres")
  )
) %>%
  layout(
    title = "Areas of parcels (in acres) moving from current zoning to ZAP",
    xaxis = list(showgrid = F, zeroline = F, title = "", showticklabels = FALSE),
    yaxis = list(showgrid = F, zeroline = F, title = "", showticklabels = FALSE)
  )

valueSankey <- plot_ly(
  type = "sankey",
  domain = list(c(x = c(0, 1), y = c(0, 1))),
  orientation = "h",
  arrangement = "snap",
  node = list(
    label = nodeswithnums$nodename,
    pad = 15, 
    thickness = 15
  ),
  link = list(
    source = flowvals$sourcenode,
    target = flowvals$targetnode,
    value = flowvals$parcelValue
  )
) %>%
  layout(
    title = "Values of parcels (in USD) moving from current zoning to ZAP",
    xaxis = list(showgrid = F, zeroline = F, title = "", showticklabels = FALSE),
    yaxis = list(showgrid = F, zeroline = F, title = "", showticklabels = FALSE)
  )

# units calcs for areas

area_sf = flowvals %>% filter(CATEGORY == "R-1 Low Density Residential") %>% pull(parcelArea) %>% sum()
area_acres = units::set_units(area_sf, "acres")

countR1 = flowvals %>% 
  group_by(
    R1 = CATEGORY == "R-1 Low Density Residential" & Zoning_2026 == "LDR") %>% 
  summarize(count = sum(parcelCount)) %>%
  mutate(percent = 100 * count / sum(count))

areaR1 = flowvals %>% 
  group_by(
    R1 = CATEGORY == "R-1 Low Density Residential" & Zoning_2026 == "LDR") %>%
  summarize(area = sum(parcelArea)) %>%
  mutate(percent = 100 * area / sum(area))

valueR1 = flowvals %>% 
  group_by(
    R1 = CATEGORY == "R-1 Low Density Residential" & Zoning_2026 == "LDR") %>%
  summarize(value = sum(parcelValue)) %>%
  mutate(percent = 100 * value / sum(value))

```

### Changes by count

```{r countsSankeyPlot, echo = FALSE}
countSankey
```

### Changes by area

```{r areaSankeyPlot, echo = FALSE}
areaSankey
```

### Changes by value

```{r valueSankeyPlot, echo = FALSE}
valueSankey
```

## Changes in total analysis

The above sankey plots shows the flow of parcels by count (the first) and by area (the second) from current zoning to the 2026 ZAP proposal. The majority of current parcels by count (`r countR1 %>% filter(R1 == TRUE) %>% pull(percent) %>% round(1)`%), and a plurality of parcels by area (`r areaR1 %>% filter(R1 == TRUE) %>% pull(percent) %>% round(1)`%) and value (`r valueR1 %>% filter(R1 == TRUE) %>% pull(percent) %>% round(1)`%), are currently and remaining the lowest density residential zoning.

The Sankey diagrams show that the majority of residential parcels that are changing zoning are moving from R-1 to MDR, which is a positive change that will allow for more residential construction and density.

### Downzoning

However!

There are 488 parcels that are inexplicably being downzoned from R-2 (analogous to MDR) to LDR. This will create more noncompliant parcels and limit the amount of housing that can be built or added in those areas.



```{r}


```

### Non-compliance

```{r nonconforming}

library(leaflet)

LDRclasses <- c("1 Family Residential", "Residential, Vacant Land", "Inn/lodge", "Det row bldg", "Aged - home")
MDRclasses <- c(LDRclasses, "2 Family Residential", "3 Family Residential", "Multiple Residential")

conformings <- parcelsJoin %>%
  mutate(
    Conforming = case_when(
      Zoning_2026 == "LDR" & CLASSDSCRP %in% LDRclasses ~ TRUE,
      Zoning_2026 == "LDR" & ! CLASSDSCRP %in% LDRclasses ~ FALSE,
      Zoning_2026 == "MDR" & CLASSDSCRP %in% MDRclasses ~ TRUE,
      Zoning_2026 == "MDR" & ! CLASSDSCRP %in% MDRclasses ~ FALSE,
      TRUE ~ NA
    ),
    ConformingCurrent = case_when(
      CATEGORY == "R-1 Low Density Residential" & CLASSDSCRP %in% LDRclasses ~ TRUE,
      CATEGORY == "R-1 Low Density Residential" & ! CLASSDSCRP %in% LDRclasses ~ FALSE,
    )
  )

nonconformingLDR <- conformings %>%
  filter(! Conforming) %>%
  filter(Zoning_2026 == "LDR") %>%
  mutate(
    popup = paste0(
      SITEADDRESS, "<br>",
      "Nonconforming because:", "<br>",
      CLASSDSCRP, " in ", Zoning_2026
    )
  )

nonconformingMap <- leaflet(nonconformingLDR %>% st_transform(4326)) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    data = zap_2026 %>% filter(Zoning_2026 == "LDR") %>% st_transform(4326),
    weight = .1, opacity = .1, color = "Grey"
  ) %>%
  addPolygons(
    weight = .2, 
    fill = TRUE,
    popup = ~popup
  ) %>%
  addLegend(
    title = "Non conforming uses in LDR",
    "topright",
    colors = c("Grey", "Blue"), labels = c("LDR zones", "Non conforming parcels")
  )

nonconformingMap

```

The above map shows LDR zones in the proposed 2026 ZAP plan, and highlights parcels that are non-conforming. I label a parcel as non-conforming in LDR if it is something other than a single family house, residential vacant land, an inn/lodge, a detached row building, or an aged home. This is an undercount because it only shows parcels that are non-conforming because of their land use class, but there are also parcels that are non-conforming because of size, setbacks, etc.

Why does it matter if a parcel is non-conforming? If an apartment building in an LDR zone is damaged by weather or fire, it cannot be rebuilt as an apartment building. Those homes will be lost and the residing families will be displaced.

What's the cause of most of the non-conforming parcels in the above map?

```{r nonconformingCases}

conformings %>% 
  filter(Zoning_2026 == "LDR") %>% 
  st_drop_geometry() %>% 
  count(Conforming, ConformingCurrent) %>%
  pivot_wider(names_from = Conforming, values_from = n) 

```

### Data & Reproducibility

All data used in this analysis is available either via the city's open data portal or the city's FOIL website. All tools in this analysis are open source. All code is posted to github. Anyone is free to reproduce and hopefully improve this analysis.

2026 ZAP zoning file released via FOIL: [download here](https://rochesterny.justfoia.com/Attachments/Download/d1d0f25a-662a-4745-963e-d2fe7797b560)

2023 ZAP zoning file also released via FOIL: [download here]()