---
title: ZAP (2026) Feedback
output:
    html_document:
        
        theme: cosmo
author: "Adrian Martin, Rochester NY"
---

## ZAP Feedback

I am writing this page as feedback to the Rochester City Planning Commission on the Zoning Alignment Project release of 2026. Feedback for the July 13th meeting is due June 18th, 4PM. This will be emailed to the commission, but also posted on github for public access and reproducibility. This is my second round of feedback on the 2026 draft; the first feedback is available [here](https://a517dogg.github.io/ZAP/index.html).

This piece will look more at non-conformities, using both the class usage description as well as lot coverage to compare ZAP's stated goal of reducing nonconformities with its actual effect (as best as I can model).

```{r setup, include=FALSE}

knitr::opts_chunk$set(cache = TRUE, echo = FALSE, warning = FALSE, message = FALSE)

library(tidyverse)
library(sf)
library(plotly)
library(leaflet)
# library(arcpullr)
library(units)

```

```{r loaddata, message = FALSE, warning = FALSE}

# note: files are transformed to a projected coordinate system
# this is necessary for accurate spatial joins, calculating areas/distances, etc.
# leaflet requires lat/lon, so I transform to 4326 after the math is done

current_zoning <- read_sf("https://maps.cityofrochester.gov/server/rest/services/Open_Data/Zoning_Districts_Open_Data/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson") %>%
    rename(Zoning_current = LABEL) %>%
    # no preservation or overlay districts
    st_transform(6541)

# parcels_2022 <- read_sf("https://maps.cityofrochester.gov/server2/rest/services/Open_Data/TaxParcel2022/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson") %>% 
#     st_transform(6541)
# API down June 8, downloaded from data.cityofrochester.gov
parcels_2022 <- read_sf("TaxParcel2022.geojson") %>% st_transform(6541)

zap_2023 <- read_sf("DRAFT_ZAP_Districts_Sept2023/DRAFT_ZAP_DISTRICTS_Sept2023.shp") %>%
    rename(Zoning_2023 = NAME) %>%
    filter( ! grepl(x = Purpose, pattern = "overlay") ) %>% 
    st_transform(6541)

zap_2026 <- read_sf("ZAP_Districts_final_draft_Apr26.shp") %>%
    rename(Zoning_2026 = NAME) %>%
    # no overlays/preservation districts in this FOIL release
    st_transform(6541)

# test = parcels_2022 %>% filter(SITEADDRESS == "1399 East Ave") %>%
#   select(BISZONING, SITEADDRESS) %>%
#   st_join(zap_2023) # overlay zones overlap with residential zones
#   # filter them out above

# for plotting
ZAPwgs <- zap_2026 %>% st_transform(4326) %>%
  # don't plot the airport overlay
  filter(Zoning_2026 != "O-A")

# footprints <- "https://maps.cityofrochester.gov/server/rest/services/Open_Data/Building_Footprint_Open_Data/FeatureServer/0" %>%
#   get_spatial_layer() %>%
#   st_transform(6541)
# above not working. just downloading instead.

footprints_all <- read_sf('data/Building_Footprints%3A_Live.geojson')
footprints <- footprints_all %>%
  # filter out the demolished structures
  filter(DEMOLISHED_TF == "Existing") %>%
  # no parcel ID; need to to a spatial join
  st_transform(6541) %>%
  # st_within doesn't work because of slight inaccuracies
  # e.g.:
  # footprintsJoin %>% filter(OBJECTID == "98325") %>% st_geometry() %>% plot()
  # arcels_2022 %>% filter(PARCELID == "12122000010550140000") %>% st_geometry() %>% plot(add = TRUE, col = "red")
  # do a centroid in parcel join instead
  mutate(centroid = st_centroid(geometry)) %>%
  st_set_geometry("centroid") %>%
  st_join(parcels_2022 %>% select(PARCELID, SITEADDRESS), st_within) %>%
  st_set_geometry("geometry") # back to footprint

# # check
# parcels_2022 %>% filter(grepl(x = SITEADDRESS, pattern = "170 Barrington")) %>% st_geometry() %>% plot()
# footprints %>% filter(grepl(x = SITEADDRESS, pattern = "170 Barrington")) %>%
#   st_geometry() %>% plot(add = TRUE)

# rules taken from page 58, and abbreviations on page 303
# https://rochesterzap.com/wp-content/uploads/2026/04/Final-Draft-Zoning-Code-April-2026.pdf

lot_coverage_rules <- tibble(
  zone = c("LDR", "MDR", "MDR", "HDR", "HDR"),
  use = list(
    "any", 
    c("SF-D", "SF-A", "2F"), c("3F", "4F", "TH", "MF", "NR"), 
    c("SF-D", "SF-A", "2F"), c("3F", "4F", "TH", "MF", "NR")
    ),
  classdscrp = list(
    "any",
    c("1 Family Residential", "1 Family Residential", "2 Family Residential"),
    # Difficult because townhouses have classdscrp of 1 family residential
    # ex: 444 Harvard, 80 Edgerton, or N Plymouth townhouses
    c("3 Family Residential", "4 Family Residential", "Townhouse", "Multiple Residential"),
    c("1 Family Residential", "1 Family Residential", "2 Family Residential"),
    c("3 Family Residential", "4 Family Residential", "Townhouse", "Multiple Residential")
  ),
  max_cov = c(50, 50, 55, 50, 55)
)

# will limit analysis to 1/2/3/4F and multiple.

# find townhouses by st_touches?

single_fams <- footprintsJoin %>% 
  filter(PARCELID %in% (parcels_2022 %>% 
    filter(CLASSDSCRP == "1 Family Residential") %>% pull(PARCELID)
    )
  )

# below gets no results with st_touches
# but takes forever to run with st_is_within_distance
# skip it!
touching <- st_is_within_distance(
  single_fams, single_fams,
  dist = 3,
  sparse = TRUE, # don't cook laptop
  remove_self = TRUE
)

# single_fams %>% 
#   mutate(hasneighbor = map_lgl(touching, ~ length(.x) > 0)) %>%
#   filter(hasneighbor)

```



```{r analysis}

zoningAreas <- current_zoning %>% 
  mutate(area = st_area(current_zoning)) %>%
  group_by(CATEGORY) %>%
  summarize(
    totalarea = sum(area),
    numDistricts = n()
  ) %>%
  st_drop_geometry() %>%
  mutate(acres= units::set_units(totalarea, "acres")) %>%
  arrange(desc(totalarea))

ZAPareas <- zap_2026 %>% 
  dplyr::filter(Zoning_2026 != "O-A") %>% #airport overlay not relevant
  mutate(area = st_area(zap_2026 %>% dplyr::filter(Zoning_2026 != "O-A"))) %>% 
  group_by(Zoning_2026) %>%
  summarize(
    totalarea = sum(area),
    numDistricts = n()
  ) %>%
  st_drop_geometry() %>%
  mutate(acres = units::set_units(totalarea, "acres")) %>%
  arrange(desc(totalarea))

# parcelsJoin <- parcels_2022 %>% 
#   mutate(parcelArea = st_area(.)) %>%
#   st_join(zap_2026, st_overlaps, largest = TRUE) %>%
#   #st_join(st_make_valid(ZAPwgs), st_overlaps) %>%
#   st_join(current_zoning, st_overlaps, largest = TRUE) %>%
#   st_join(zap_2023, st_overlaps, largest = TRUE)

#   # above takes forever; let's just run it once and reload it.

# saveRDS(parcelsJoin, "parcelsJoin.RDS")
parcelsJoin <- readRDS("parcelsJoin.RDS")

footprintsJoin <- footprints %>%
  select(
    OBJECTID, GEOMETRYSOURCE, BUILDINGCLASS, PARCELID, SITEADDRESS,
    geometry, centroid) %>%
  # sample_n(100) %>%
  left_join(
    parcelsJoin %>% 
      as_data_frame() %>%
      select(PARCELID, CLASSDSCRP, Zoning_current, Zoning_2026, parcelgeo = geometry),
    by = "PARCELID"
  ) %>%
  mutate(
    lotCoveragePercent = as.numeric(100 * st_area(geometry) / st_area(parcelgeo))
  ) %>% 
  left_join(
    lot_coverage_rules %>% unnest(cols = c(use)) %>% unnest(cols = c(classdscrp)), 
    by = c("Zoning_2026" = "zone", "CLASSDSCRP" = "classdscrp")
  ) %>%
  mutate(
    max_cov = ifelse(Zoning_2026 == "LDR", .5, max_cov))
  

# multiple footprints for some parcels (garages etc)

<!-- footprints_by_parcel <- footprintsJoin %>%
  group_by(PARCELID) %>%
  summarize(
    total_lot_cov = sum(lotCoveragePercent)
  ) %>%
  left_join(
    footprintsJoin %>% 
      st_drop_geometry() %>%
      select(PARCELID, SITEADDRESS, CLASSDSCRP, Zoning_2026, max_cov),
      by = "PARCELID"
  ) %>%
  # now to do the judgement on conforming or not
  mutate(
    conforming_cov = total_lot_cov < max_cov
  ) -->

# that takes a long time on my shitty computer
# saveRDS(footprints_by_parcel, "footprints_by_parcel.RDS")
footprints_by_parcel <- readRDS("footprints_by_parcel.RDS")

parcelsSummed <- parcelsJoin %>% 
  st_drop_geometry() %>%
  group_by(CATEGORY, Zoning_2026) %>% 
  summarize(
    parcelCount = n(), 
    parcelArea = sum(parcelArea),
    parcelValue = sum(CURRENT_TOTAL_VALUE, na.rm = TRUE)) # 59 parcels with NA value

# footprint join

# build logic for what parcels are allowed what lot coverage

# check for lot coverage per parcel

```


```{r, echo = FALSE}

downzoned <- parcelsJoin %>%
  filter(
    (Zoning_2026 == "LDR" & CATEGORY != "R-1 Low Density Residential") |
      (Zoning_2026 == "MDR" & CATEGORY != "R-1 Low Density Residential" & CATEGORY != "R-2 Medium Density Residential" & CATEGORY != "URD Urban Renewal District")
  ) %>%
  mutate(
    NewlyNonConforming = case_when(
      (Zoning_2026 == "LDR" & CLASSDSCRP %in% c("1 Family Res", "Res vac land", "Inn/lodge", "Det row bldg")) ~ FALSE,
      (Zoning_2026 == "MDR" & CLASSDSCRP %in% c("Apartment", "Animal welfr")) ~ TRUE,
      Zoning_2026 == "MDR" ~ FALSE,
      TRUE ~ TRUE
    ),
    popup = paste(
      SITEADDRESS, "<br>",
      "Current: ", CATEGORY, "<br>",
      "ZAP: ", Zoning_2026, "<br>",
      ifelse(NewlyNonConforming, "Newly nonconforming", "Still conforms")
    ),
    fill = case_when(
      Zoning_2026 == "LDR" ~ "Blue",
      Zoning_2026 == "MDR" ~ "Orange"
    ),
    weight = case_when(
      NewlyNonConforming == TRUE ~ .5,
      NewlyNonConforming == FALSE ~ .1
    )
  )

downzoneMap <- leaflet(downzoned %>% st_transform(4326)) %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    popup = ~popup, 
    weight = ~weight,
    color = ~fill, fill = TRUE) %>%
  addLegend(
    "topright",
    title = "Areas downzoned by ZAP",
    colors = c("Orange", "Blue"), labels = c("Downzoned to MDR", "Downzoned to LDR")
  )

# downzoneMap

```

### Non-compliance

```{r nonconforming}

R1_parcels = parcels_2022 %>% 
  st_drop_geometry() %>%
  filter(BISZONING == "R-1") %>%
  count(CLASSDSCRP) %>%
  mutate(percent = 100 * n / sum(n))

LDRclasses <- c("1 Family Residential", "Residential, Vacant Land", "Det row bldg", "Aged - home")
MDRclasses <- c(LDRclasses, "2 Family Residential", "3 Family Residential", "Multiple Residential")

conformings <- parcelsJoin %>%
  mutate(
    Conforming = case_when(
      Zoning_2026 == "LDR" & CLASSDSCRP %in% LDRclasses ~ TRUE,
      Zoning_2026 == "LDR" & ! CLASSDSCRP %in% LDRclasses ~ FALSE,
      Zoning_2026 == "MDR" & CLASSDSCRP %in% MDRclasses ~ TRUE,
      Zoning_2026 == "MDR" & ! CLASSDSCRP %in% MDRclasses ~ FALSE,
      TRUE ~ NA
    ),
    ConformingCurrent = case_when(
      CATEGORY == "R-1 Low Density Residential" & CLASSDSCRP %in% LDRclasses ~ TRUE,
      CATEGORY == "R-1 Low Density Residential" & ! CLASSDSCRP %in% LDRclasses ~ FALSE,
    )
  )

nonconformingLDR <- conformings %>%
  filter(! Conforming) %>%
  filter(Zoning_2026 == "LDR") %>%
  mutate(
    popup = paste0(
      SITEADDRESS, "<br>",
      "Nonconforming because:", "<br>",
      CLASSDSCRP, " in ", Zoning_2026
    )
  )

nonconformingMap <- leaflet(nonconformingLDR %>% st_transform(4326)) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    data = zap_2026 %>% filter(Zoning_2026 == "LDR") %>% st_transform(4326),
    weight = .1, opacity = .1, color = "Grey"
  ) %>%
  addPolygons(
    weight = .2, 
    fill = TRUE,
    popup = ~popup
  ) %>%
  addLegend(
    title = "Non conforming uses in LDR",
    "topright",
    colors = c("Grey", "Blue"), labels = c("LDR zones", "Non conforming parcels")
  )

nonconformingMap

```

The above map shows LDR zones in the proposed 2026 ZAP plan, and highlights parcels that are non-conforming based on the parcel description. I label a parcel as non-conforming in LDR if it is something other than a single family house, residential vacant land, an aged home, or a detached row building (likely a townhouse which is allowed in LDR, a very welcome change from the 2023 proposal). This is an *undercount* because it only shows parcels that are non-conforming because of their land use class, but there are also parcels that are non-conforming because of size, setbacks, etc. Had I more time with the shapefile, I could have forced city staffers to read even more analysis that grouped parcels by street segment, calculated average lot frontages per block, and estimated non-conforming parcels by building and parcel dimension.

Why does it matter if a parcel is non-conforming? If an apartment building in an LDR zone is damaged by weather or fire, it cannot be rebuilt as an apartment buildingl those homes will be lost and the residing families will be displaced. If a family lives in one of the 5,364 duplexes in an LDR zone, they're legally prohibited from building an addition (or ADU) since it would be classified as an increase in intensity.

One of the stated purposes of this zoning code is to gradually eliminate non-conforming uses and reuse or eliminate non-conforming structures. Thousands of our neighbors (including me!) live in non-conforming residential structures because the code is too strict.

What's the cause of most of the non-conforming labels in the above map?

```{r nonconformingCases}

nonconformingLDR %>%
  st_drop_geometry() %>% 
  group_by(CLASSDSCRP) %>%
  count() %>%
  mutate(Description = case_when(
    n > 50 ~ CLASSDSCRP, TRUE ~ "Other")
  ) %>%
  group_by(Description) %>%
  summarize(count = sum(n)) %>%
  mutate(percent = 100 * count / sum(count)) %>%
  arrange(desc(count)) %>%
  knitr::kable()

```

The vast majority (over 80%) of non-conforming parcels in LDR are duplexes. Over 14% of all parcels in LDR are duplexes. Duplexes should be encouraged in LDR.

Appartments are only 265 of the nonconforming parcels, but those 265 parcels contain (according to 2022 parcel data) `r nonconformingLDR %>% filter(CLASSDSCRP == "Apartment") %>% pull(NORESUNITS) %>% sum()` residential units. When we plot them on a map, it's clear that apartments in LDR are clustered: in the southeast, in Swillburg, Upper Monroe and Park Ave, and in the northwest between Dewey and Lake, north of Driving Park. These areas should probably be rezoned to MDR or one of the neighborhood center classifications, however that opportunity was apparently lost with the Roc2034 adoption. The possibility of those neighborhoods losing housing due to normal attrition and aging of apartment buildings should concern the city as well as businesses in those neighborhoods, who would be losing customers. The city [acknowledges in responses to ZAP comments](https://rochesterzap.com/wp-content/uploads/2026/05/FGEIS-Comment-Response-Summary-Table.pdf) that the density in Park Ave contributes to the success of businesses in the area.^["More tall buildings with residents offers more customers for local businesses. The few tall apartment buildings along Park Ave contribute to the success of businesses along that corridor." page 16] An easy way to address this is to allow new duplexes in LDR, which would allow LDR to be more flexible to responses in residential demand.

### Changes from 2023 to 2026

```{r changes2023to2026}

changes_from_draft <- parcelsJoin %>%
  filter(Zoning_2026 != Zoning_2023) %>%
  select(SITEADDRESS, BISZONING, Zoning_2023, Zoning_2026) %>%
  mutate(
    popup = paste0(
      SITEADDRESS, " is ", BISZONING, "<br>",
      "2023: ", Zoning_2023, "<br>",
      "2026: ", Zoning_2026
      )
    ) %>%
  st_transform(4326)

leaflet(changes_from_draft) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(weight = .2, opacity = .2, color = "Red", popup = ~popup)


```

## Conclusion





### Data & Reproducibility

All data used in this analysis is available either via the city's open data portal or the city's FOIL website. All tools in this analysis are open source. All code is posted to [github](https://github.com/a517dogg/ZAP). Anyone is free to reproduce and hopefully improve this analysis.

2026 ZAP zoning file released via FOIL: [download here](https://rochesterny.justfoia.com/Attachments/Download/d1d0f25a-662a-4745-963e-d2fe7797b560)

2023 ZAP zoning file also released via FOIL: [download here](https://github.com/a517dogg/ZAP/blob/main/DRAFT_ZAP_Districts_Sept2023.zip)